cardinalR: Generating interesting high-dimensional data structures

### 2025-09-23 paper-cardinalR.Rmd
Updated abstract with all the comments
Removed the installation section

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. These datasets contain nonlinear manifolds in image and speech recognition, clusters in genomics and forensic analysis, and sparse distributions in text mining. We can generate data containing a variety of structures using mathematical functions and statistical distributions. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples, adding to the existing toolset of benchmark datasets for evaluating algorithms.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

## Installation

The development version can be installed from GitHub:

```r
pak::pak("JayaniLakshika/cardinalR")
```

>>>>

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. These datasets contain nonlinear manifolds in image and speech recognition, clusters in genomics and forensic analysis, and sparse distributions in text mining. Data with a variety of structures can be generated using mathematical functions and statistical distributions to create test datasets. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples, adding to the existing toolset of benchmark datasets for evaluating algorithms.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

The `cardinalR` R package is available on GitHub at [JayaniLakshika/cardinalR](https://github.com/JayaniLakshika/cardinalR).
